I am compiling the Texas House election return data.

```{r}
#| label: setup
#| message: FALSE

library(tidyverse)
library(janitor)
```

I am creating a list of the general election results.

```{r}
general_returns <- list.files(
  "data-raw/house election returns",
  pattern = "General",
  full.names = TRUE)
```

I am importing the general election return files.

```{r}
general_raw <- general_returns |> 
  set_names(basename) |> 
  map(read_csv) |> 
  list_rbind(names_to = "year") |>
  clean_names()

general_raw
```

I am cleaning the general returns file to include just Texas House results.

```{r}
general_clean <- general_raw |> 
  mutate(year = str_remove(year, "_General_Election_Returns.csv")) |> 
  filter(grepl("State Rep", office, ignore.case = TRUE)) |> 
  mutate(district = str_remove(office, "State Rep")) |>
  select(year, cntyvtd, district, name, party, incumbent, votes) |> 
  group_by(year, district, name) |> 
  summarize(votes = sum(votes)) |> 
  mutate(district = as.numeric(district)) |> 
  arrange(district)

general_clean
```

I am exporting the general returns.

```{r}
general_clean |> 
  write_rds("data-clean/01-general-clean.rds")
```


I am creating a list of the primary election results, not including runoffs.

```{r}
primary_returns <- list.files(
  "data-raw/house election returns",
  pattern = "Primary",
  full.names = TRUE)
```

I am importing the primary election return files.

```{r}
primary_raw <- primary_returns |> 
  set_names(basename) |> 
  map(read_csv) |> 
  list_rbind(names_to = "year") |>
  clean_names()
```

I am cleaning the primary return files and filtering to just State House races.

```{r}
primary_clean <- primary_raw |> 
  mutate(year = str_remove(year, "_Democratic_Primary_Election_Returns.csv")) |> 
  mutate(year = str_remove(year, "_Republican_Primary_Election_Returns.csv")) |> 
  filter(grepl("State Rep", office, ignore.case = TRUE)) |> 
  mutate(district = str_remove(office, "State Rep")) |> 
  select(year, cntyvtd, district, name, party, incumbent, votes) |> 
  group_by(year, district, name) |> 
  summarize(votes = sum(votes)) |> 
  mutate(district = as.numeric(district)) |> 
  arrange(district)

primary_clean
```

I am exporting the primary returns.

```{r}
primary_clean |> 
  write_rds("data-clean/01-primary-clean.rds")
```

