---
title: "Combine results"
author: "Media Innovation Group"
code-fold: true
code-summary: "Expand this to see code"
---

This notebook compiles Voting Tabulation Districts election returns from different years, processes them into state-wide results, then filters them for State Rep.

Originally created by MIG data fellow Isabella Zeff, it has since been refactored by Christian McDonald.

The data some from the Texas Legislative Council's [data portal](https://data.capitol.texas.gov/dataset/comprehensive-election-datasets-compressed-format). The [documentation is available here](https://data.capitol.texas.gov/dataset/vtds). We used the 2024 General VTDs Election Data CSV version that includes 2012 - 2024 election data reported by 2024 primary election VTDs.

## Setup

```{r}
#| label: setup
#| message: FALSE

library(tidyverse)
library(janitor)
```

## Functions

Function to create totals from the county-by-county results

```{r}
fun_totals_all <- function(.data){
  .data |> 
    group_by(year, election, office, name, party, incumbent) |> 
    summarize(candvotes = sum(votes), .groups = "drop") |> 
    arrange(year, office, candvotes |> desc())
}
```

## Import and combine

```{r}
all_files <- list.files(
  "data-original/vdt-returns",
  pattern = ".csv",
  full.names = TRUE)
```

```{r}
#| message: false
#| warning: false

all_raw <- all_files |> 
  set_names(basename) |> 
  map(\(x) read_csv(x, col_types = cols(.default = col_character()))) |> 
  list_rbind(names_to = "source") |>
  clean_names()
```

## Clean source

Here we use the name of the file to find the election year and name. We also turn votes into a number.

```{r}
all_returns <- all_raw |> 
  mutate(
    year = str_sub(source, 1, 4),
    election = str_sub(source, 6, -22) |> str_replace_all("_", " "),
    .before = county
  ) |> 
  mutate(votes = votes |> as.numeric()) |> 
  select(-source)

all_returns |> head()
```

## Tally votes

```{r}
all_totals <- all_returns |> 
  fun_totals_all() |> 
  arrange(year, election, office, candvotes |> desc())

all_totals |> head()
```

### Filter for State Reps

Find the state reps

```{r}
rep_totals <- all_totals |> 
  filter(str_detect(office, "State Rep")) |> 
  mutate(district = parse_number(office), .after = office)
```

## Export results

I am exporting just the rep results at this point..

```{r}
all_totals |> 
  write_rds("data-processed/01-all-totals.rds")

rep_totals |> 
  write_rds("data-processed/01-rep-totals.rds")
```

