---
title: "03-district-results"
format: html
---

> A possible flaw in the original data is we don't account for runoffs in primaries (or general, but would there be? this is shown with 2016 Dist 27 Dem Primary.)

Beyond lack of filerID, the big difference in what we have now in our competitive races file vs what we need is the current data is by election .. .i.e., a different row for democratic primary vs gop primary, etc.

We need to reconfigure that so there is one row for each year/district, and then a number of variables from each type of election (both primaries, both primary runoffs, general, general runoffs) with the winner’s last name (all we have), incumbency!, and the winning margin since we have it.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false

library(tidyverse)
library(janitor)
```

## Import

```{r}
elcts <- c(
  "Democratic Primary",
  "Democratic Runoff",
  "Republican Primary",
  "Republican Runoff",
  "General"
)

totals_rep <- read_rds("data-processed/01-rep-totals.rds")

margins_raw <- read_rds("data-processed/02-reps-margins.rds")

margins_raw |> count(election)

margins_reorder <- margins_raw |> 
  filter(election %in% elcts,
         year >= 2016) |> 
  mutate(election = fct(election, levels = elcts)) |> 
  relocate(district, .after = year) |> 
  arrange(year, district, election)

# this is a breaking change in i'm testing a single year
margins_reorder <- margins_reorder |> filter(year == 2024)
```

## Methods discussions

- I could filter for each election type individually, the reorganize those.

```{r}
dp <- margins_reorder |> 
  filter(election == "Democratic Primary") |> 
  select(
    year, district,
    dp_win = winner,
    dp_inc = winner_inc,
    dp_mrg = winning_margin
  )

dr <- margins_reorder |> 
  filter(election == "Democratic Runoff") |> 
  select(
    year, district,
    dr_win = winner,
    dr_inc = winner_inc,
    dr_mrg = winning_margin
  )

rp <- margins_reorder |> 
  filter(election == "Republican Primary") |> 
  select(
    year, district,
    rp_win = winner,
    rp_inc = winner_inc,
    rp_mrg = winning_margin
  )

rr <- margins_reorder |> 
  filter(election == "Republican Runoff") |> 
  select(
    year, district,
    rr_win = winner,
    rr_inc = winner_inc,
    rr_mrg = winning_margin
  )

gen <- margins_reorder |> 
  filter(election == "General") |> 
  select(
    year, district,
    g_win = winner,
    g_inc = winner_inc,
    g_mrg = winning_margin
  )


combo <- 
  dp |> 
  full_join(dr, by = join_by(year, district)) |> 
  full_join(rp, by = join_by(year, district)) |> 
  full_join(rr, by = join_by(year, district)) |> 
  full_join(gen, by = join_by(year, district))
  

combo |> arrange(year, district)
```


## data smells

Why so many NAs? We can look to  [Ballotpedia in 2024](https://ballotpedia.org/Texas_House_of_Representatives_elections,_2024#Primary) to see why

### District 1

- Democratic Primary: No candidates; no primary.
- Republican Primary: Gary Vandeaver won.
- General: Vandeaver was unopposed, so no results.

### District 2

- Democratic Primary: There was an unopposed Democrat, so no primary.
- Republican Primary: Brent Money won.
- General: Money was unopposed, so no results.

### District 3

_So there are NO results for District 3 in 2024_

- Democratic Primary: No candidates; no primary.
- Republican Primary: Cecil Bell was unopposed, so no results.
- General: Cecil Bell was unopposed, so no results.

### District 4

- Democratic Primary: Alex Bar-Sela unopposed, so no results
- Republican Primary: Keith Bell won primary.
- General: Bell won the general.

```{r}
margins_raw |> 
  filter(year == 2024, district == 3)
```

```{r}
totals_rep |> 
  filter(year == 2024, district == 3)
```








